%\documentclass{...}

\input{./def/yf-formatting}

\usepackage{amsfonts, amsmath, amssymb, amsthm}
\usepackage{comment}
\usepackage{graphicx}
\usepackage{ifthen}
\usepackage{latexsym}
%\usepackage{times}
\usepackage[normalem]{ulem}

\input{./def/yf-def}

\def\agm{\text{AGM}}
\def\B{\mathcal{B}}
\def\E{\mathcal{E}}
\def\H{\mathcal{H}}
\def\Q{\mathcal{Q}}
\def\V{\mathcal{V}}
\def\W{\mathscr{W}}

\newboolean{solver}\setboolean{solver}{true}
%\newboolean{solver}\setboolean{solver}{false}
\ifthenelse{\boolean{solver}}{\includecomment{sol}}{\excludecomment{sol}}

\begin{document}

\section*{CE7456: Final Exam}

\noindent Name:  \hspace{50mm}
Student ID:

% \begin{center}
%     \uline{Classroom Discussion}
% \end{center}

\extraspacing

\fbox{
    You do not need to describe any algorithm in our lecture notes.
}

\extraspacing {\bf Problem XX (YY marks).} Set $S = \set{e_1, e_2, ..., e_5}$ where each element carries a positive integer weight. The following describes an alias structure on $S$:
\myitems{
    \item Urn 1: elements $e_1$ and $e_2$, associated with values $10/5$ and $2/5$, respectively;
    \item Urn 2: elements $e_2$ and $e_4$, associated with values $8/5$ and $4/5$, respectively;
    \item Urn 3: element $e_3$, associated with value $12/5$;
    \item Urn 4: elements $e_1$ and $e_5$, associated with values $1/5$ and $11/5$, respectively;
    \item Urn 2: elements $e_2$ and $e_4$, associated with values $8/5$ and $4/5$, respectively.
}

\extraspacing {\bf Problem XX (YY marks).} A {\em 3-star} is the graph $P = (V_P, E_P)$ where $V_P = \set{u, v_1, v_2, v_3}$ and $E_P = \set{(u, v_1), (u, v_2), (u, v_3)}$. From the AGM bound, an undirected graph of $m$ edges can have at most $O(m^3)$ occurrences of $P$. Construct such a graph $G = (V, E)$ of $m$ edges that have $\Omega(m^3)$ occurrences of $P$.

\extraspacing {\bf Problem XX (YY marks).} Given a set $S$ of $n$ integers, the goal of the {\em 3-sum problem} is to decide whether there are three distinct elements $x, y, z \in S$ such that $x + y + z = 0$. Answer the following questions:
\myitems{
    \item Design an MPC algorithm that solves the problem in constant rounds with load $O(n / \sqrt{p})$, where $p$ is the number of machines. You may assume that each machine stores $n / p$ elements in the beginning except possibly the last machine.

    \item Design an EM algorithm that solves the problem in $O(n^2 / (MB))$ I/Os, where $M$ is the number of words in memory, and $B$ is the number of words in a disk block. You may assume that the set $S$ is given in a disk array of $n$ words.

    \vgap

    Note: You get 70\% of the marks if your algorithm performs $\Theta(\fr{n^2}{MB} \polylog n)$ I/Os.
}

\extraspacing {\bf Problem XX (YY marks).} Let $\Q$ be a join whose schema graph $\H = (\V, \E)$ has a constant size. Take any fractional edge cover $\W$ of $\H$. Prove: if the AGM value of $\Q$ under $\W$ is 1, then for every attribute $A \in \V$, there is a relation $R \in \Q$ satisfying
\myitems{
    \item $A$ belongs to the schema of $R$;
    \item $|R| = 1$.
}

\extraspacing {\bf Problem XX (YY marks).} Let $R_1, R_2$, $R_3$, and $R_4$ be relations whose schemas are $\set{A, B, C}$, $\set{A, C, D}$, $\set{B, D}$, and $\set{B, C, E}$, respectively. Consider the natural join $\Q = \set{R_1, R_2, R_3, R_4}$. Suppose that we insert a tuple $\bm{u}$ in $R_4$ --- let us denote the new relation as $R_4'$, namely, $R_4' = R_4 \cup \set{\bm{u}}$. Consider the natural join $\Q' = \set{R_1, R_2, R_3, R_4'}$.

\vgap

Prove: the result of $\Q'$ can have at most $\sqrt{|R_1| \cdot |R_2| \cdot |R_3|}$ tuples that are not in the result of $\Q$.


\end{document}
