
\include{./def/yf-formatting}
\include{./def/yf-def}

\usepackage{wasysym}
%\usepackage{times}

\title{}
\author{}
\date{}


%\def\A{\mathcal{A}}
%\def\B{\mathcal{B}}
\def\C{\mathcal{C}}
%\def\D{\mathcal{D}}
\def\E{\mathcal{E}}
\def\FF{\EuScript{F}}
\def\G{\mathcal{G}}
\def\GG{\EuScript{G}}
\def\H{\mathcal{H}}
\def\I{\mathcal{I}}
\def\J{\mathcal{J}}
\def\L{\mathcal{L}}
\def\Q{\mathcal{Q}}
\def\S{\mathcal{S}}
\def\T{\mathcal{T}}
\def\V{\mathcal{V}}
%\def\W{\EuScript{W}}
\def\W{\mathscr{W}}
\def\X{\mathcal{X}}
\def\Y{\mathcal{Y}}
\def\Z{\mathcal{Z}}
\def\att{\mit{diffatt}}
\def\agm{\mathit{AGM}}
\def\boldatt{\mathbf{att}}
\def\ext{\mathit{ext}}
\def\In{\mathrm{IN}}
\def\join{\mathit{join}}
\def\Out{\mathrm{OUT}}
\def\schema{\mathit{schema}}

\def\extraspacing{\vspace{4mm} \noindent}
\def\vgap{\vspace{4mm}}
\def\vslit{\vspace{2mm}}

\begin{document}

\boxminipg{\linewidth}{
    \begin{center}
        \vspace{3mm}
    {\Large
    (CE7456 Lecture Notes 14) Massively Parallel Computation} \\[2mm]
    {\large Yufei Tao}
    \vspace{3mm}
    \end{center}
}

We have seen two computational models so far: (i) the RAM model, which is suitable when the data fit in main memory, and (ii) the EM model, which is suitable when the data fit on the disk (of one machine). In practice, many applications need to process datasets so large that they must be stored on multiple machines. The performance bottleneck in processing such data --- e.g., in systems like Hadoop and Spark --- is neither CPU computation nor I/O access; rather, it is typically the communication across machines. The RAM and EM models are no longer appropriate for studying algorithms in such scenarios.

\vgap

This lecture introduces the {\em massively parallel computation} (MPC) model \cite{bks17,ksv10} for studying algorithms designed to minimize parallel communication over a large number of machines. We will then illustrate the characteristics of the model through several representative algorithms.

\extraspacing {\bf Math Conventions.} Given an integer $x \ge 1$, we use $[x]$ to represent the set $\set{1, 2, ..., x}$.

\section{The MPC Model} \label{sec:model}

In this model, we have $p$ share-nothing machines that are interconnected in a network. Denote by $S$ the input set of elements, and let $n = |S|$. In the beginning, each machine stores at most $n/p$ elements of $S$; an algorithm has no control over how elements are distributed across the machines. As a standard assumption, the values of $p$ and $n$ satisfy $p < n^{1-\eps}$, where $\eps$ is a positive constant less than 1.


\vgap

An algorithm starts by having each machine perform some initial computation on its local data. Then, it executes in {\em rounds}, each having two phases:
\myitems{
  \item in the first phase, the machines exchange messages (every message should have been prepared either in the initial computation or the second phase of the previous round);
  \item in the second phase, each machine performs local computation.
}
An algorithm can perform any number of rounds as needed. Upon termination, the final output must be available on the machines in the form specified by the problem definition.

\vgap

The {\em load of a round} is the largest number of words communicated (sent and received combined) by a machine in that round; specifically, if machine $i \in [p]$ communicates $\ell_i$ words in this round, then the load of the round is $\max_{i=1}^p \ell_i$.
The (total) {\em load} of the algorithm is the sum of the loads of all the rounds.

\vgap

With load $n$, any problem can be solved trivially in one round by simply sending all data to one machine. In practice, a good algorithm ought to perform a small number of rounds and incur a small load. Accordingly, two interesting questions arise in theory:
\myitems{
    \item ({\bf Load complexity}) What is the smallest load for solving a problem within a given number of rounds?

    \item ({\bf Round complexity}) What is the smallest number of rounds for solving a problem when each round can incur no more than a given load?
}

\section{Sorting} \label{sec:sort}

The sorting problem is defined as follows in the MPC model. Let $S$ be a set of $n$ integers, initially distributed across $p$ machines in an arbitrary manner, subject to the constraint that each machine has at most $n/p$ elements. The goal is to re-distribute the elements across the machines such that the elements on machine $i$ are less than those on machine $j$, for any $1 \le i < j \le p$.

\vgap

Assuming $n \ge 30$ and $n \ge p^2 \ln n$, we show how to solve the problem in 2 rounds, achieving a load of $O(n/p)$ with probability at least $1 - 1/n^2$.

\extraspacing {\bf Algorithm.} For each $i \in [p]$, denote by $S_i$ the set of elements stored on machine $i$ initially. Set
\myeqn{
    \pi = (p / n) \cdot \ln n
}
In the first phase of round 1, each machine generates a sample set $\Sigma_i \subseteq S_i$ as follows: for each element of $S_i$, independently toss a coin with probability $\pi$ of heads and add the element to $\Sigma_i$ if the coin comes up heads.

\vgap

In the second phase of round 1, machine $i \in [p]$ sends $\Sigma_i$ to every other machine.

\vgap

In the first phase of round 2, each machine collects
\myeqn{
    S_\mit{sam} &=& \bigcup_{j=1}^p \Sigma_j. \label{eqn:sort:S_sam}
}
Set
\myeqn{
    s = |S_\mit{sam}|. \label{eqn:sort:s}
}
The machine identifies $p - 1$ elements $b_1, b_2, ..., b_{p-1}$ where $b_j$ ($1 \le j \le p-1$) is the $j \cdot \lc s/p \rc$-th smallest element in $S_\mit{sam}$. In addition, define $b_0 = -\infty$ and $b_p = \infty$. We refer to $b_0, b_1, ..., b_p$ the as {\em boundary elements}.

\vgap

In the second phase of round 2, machine $i \in [p]$ sends to machine $j \in [p]$ the elements in $S_i \cap (b_{i-1}, b_i]$.

\extraspacing {\bf Analysis.}
The algorithm incurs a load of $O(\fr{n}{p} + p \ln n) = O(n/p)$ when the following two conditions are met (think: why?):
\myitems{
    \item {{\bf C1:}} $|\Sigma_i| = O(\ln n)$ for all $i \in [p]$.
    %\item {{\bf C2:}} $s = O(m)$.
	\item {{\bf C2:}} $S \cap (b_{i-1}, b_i]$ has $O(n/p)$ elements for all $i \in [p]$.
}

\begin{theorem} \label{thm:sort}
	When $n \ge \max\{30, p \ln n\}$, {\bf C1} and {\bf C2} hold simultaneously with probability at least $1 - 1/n^2$.
\end{theorem}

\begin{proof}
	Consider an arbitrary $i \in [p]$. Define $X_i = |\Sigma_i|$. Note that $X_i$ is a random variable with $\expt[X_i] \le (n/p) \cdot \pi = \ln n$. An application of Chernoff bound \eqref{eqn:chernoff2} gives:
	\begin{eqnarray}
		\Pr[X_i \ge 18 \ln n] \le \exp(- 3 \ln n) = 1/n^3. \nonumber
	\end{eqnarray}
	By union bound, the probability that {\bf C1} fails to hold --- namely,
	$X_i \ge 18 \ln n$ for at least one $i \in [p]$ --- is at most $p/n^3$.

	\vgap

	Next, we consider {\bf C2}. First, by the above analysis, we can assert
	\myeqn{
        \Pr[s \ge 18 p \cdot \ln n] \le p/n^3 \nn
	}
	where the value of $s$ is given in \eqref{eqn:sort:s}.
	Unless otherwise stated, the subsequent discussion assumes $s < 18 p \cdot \ln n$.

	\vgap

	Imagine that $S$ has been sorted in ascending order. Let us  partition the sorted list into $\lc p/64 \rc$ sub-lists --- call each sub-list a {\em bucket} --- such that each bucket has between $64 n/p$ and $128 n/p$ objects (think: why is it fair to assume such a partitioning?). Here is an important observation:

	\begin{center}
        If every bucket covers at least one boundary element, {\bf C2} holds.
	\end{center}

	\noindent Indeed, under the above if-condition, no bucket can fall between two consecutive boundary elements; hence, for each $i \in [t]$, the set $S \cap (b_{i-1}, b_i]$ can contain elements from at most 2 buckets, implying $S \cap (b_{i-1}, b_i]$ has size at most $256 n/p$.

	\vgap

	Fix an arbitrary bucket $S_\mit{buk}$. Under the event $s < 18p \cdot \ln n$, the bucket must include a boundary element if it has at least $18 \ln n$ --- which is greater than $s/p$ --- elements from $S_\mit{sam}$ (as defined in \eqref{eqn:sort:S_sam}) because a boundary element is taken every $\lc s/p \rc$ consecutive elements in the sorted order of $S_\mit{sam}$.

	\vgap

	For each $j \in [|S_\mit{buk}|]$, define random variable $Y_j = 1$ if the $j$-th element of $S_\mit{buk}$ is sampled in the first phase of round 1, or 0 otherwise. Then,
	\myeqn{
        Y &=& \sum_{j=1}^{|S_\mit{buk}|} Y_j. \nn
	}
	is the number of elements of $S_\mit{sam}$ that are contained in $S_\mit{buk}$. Clearly, $\expt[Y] \ge (64n / p) \cdot \pi = 64 \ln n$.

	\vgap

	We can derive:
	\begin{eqnarray}
		\Pr[Y < 18 \ln n] &=& \Pr[Y \le (1 - 46/64) \cdot 64 \ln n] \nn \\
		&\le& \Pr[Y \le (1 - 46/64)\expt[Y]] \nn \\
		\textrm{(by Chernoff bound \eqref{eqn:chernoff1})} &\le& \exp\left(- \fr{46^2}{64^2} \cdot \fr{\expt[Y]}{3}\right) \nn \\
		&\le& \exp\left(- \fr{46^2}{64^2} \cdot \fr{64 \ln n}{3}\right) \nn \\
		&<& \exp(-3 \ln n) \nn \\
		&=& 1/n^3. \nn
	\end{eqnarray}
	Therefore, the probability that $S_\mit{buk}$ does not cover any boundary element is at most $1/n^3$.

	\vgap

	As there are $\lc p/64 \rc \le p$ buckets, the probability that at least one bucket fails to cover any boundary element is at most $p/n^3$. We can now conclude that {\bf C2} fails to hold with probability at most $2p/n^3$ (think: why does the constant 2 come from?).

	\vgap

	Therefore, {\bf C1} and {\bf C2} hold at the same time with probability at least $1 - 3p/n^3$, which is at least $1 - 1/n^2$ because $n \ge 3p$.
\end{proof}

\extraspacing {\bf A Lower Bound.} For any values of $p$ and $n$, any algorithm must incur a load of $\Omega(n/p)$ in the worst case, no matter how many rounds are performed (think: why?).

\section{Cartesian Product} \label{sec:cp}

The {\em Cartesian product} (CP) problem is defined as follows in the MPC model. Let $S_1, S_2, ..., S_t$ be $t$ sets of elements, where $t$ is a constant integer at least 2. Let $n = \sum_{i=1}^t |S_i|$. The elements of $S_1, ..., S_t$ are initially distributed across $p$ machines in an arbitrary manner, subject to the constraint that each machine has at most $n/p$ elements. The goal is to produce each $t$-tuple $(u_1, u_2, ..., u_t) \in S_1 \times S_2 \times ... \times S_t$ on at least one machine.

\vgap

Next, we will show how to solve the problem in one round with load $O(n / p^{1/t})$. Our description makes the following assumption: for every $S_i$ ($i \in [t]$), each element of $S_i$ is associated with a unique ID from $[|S_i|]$.

\extraspacing {\bf Algorithm.} Define
\myeqn{
    p^* = \lf p^{1/t} \rf^t. \nn
}
Note that $p^* > (p^{1/t} / 2)^t = p^t / 2^t = \Omega(p^t)$.

\vgap

Conceptually, align machines 1, 2, ..., $p^*$ into a $t$-dimensional
\myeqn{
    \underbrace{\lf p^{1/t} \rf \times \lf p^{1/t} \rf \times ... \times \lf p^{1/t} \rf}_t \nn
}
grid. Each machine in the grid can be labeled as a $t$-dimensional point $(x_1, x_2, ..., x_t)$ where $1 \le x_i \le \lf p^{1/t} \rf$ for all $i \in [t]$. We will refer to each $x_i$  as the machine's {\em coordinate} on dimension $i$.

\vgap

Each machine examines the elements in its (initial) local storage as follows. Let $e$ be such an element; assume that $e$ originates from $S_i$ (for some $i \in [t]$) and carries an ID $y$. The machine sends $e$ to every machine in the grid whose coordinate on dimension $i$ equals
\myeqn{
    1 + \left(y \mod \lf p^{1/t} \rf \right). \label{eqn:cp:coord-on-i}
}

\extraspacing {\bf Correctness.} We prove that each $(u_1, u_2, ..., u_t) \in S_1 \times S_2 \times ... \times S_t$ is produced on at least one machine.

\vgap

For each $i \in [t]$, let $y_i$ be the ID of tuple $u_i \in S_i$, and define
\myeqn{
    x_i = 1 + \left(y_i \mod \lf p^{1/t} \rf \right). \nn
}
Thus, the machine with coordinates $(x_1, x_2, ..., x_t)$ definitely receives all the $t$ tuples $u_1, u_2, ..., u_t$.

\extraspacing {\bf Load Analysis.} Let us first discuss how many elements are sent by a machine. As mentioned, the machine sends an element $e \in S_i$ with ID $y$ to all machines whose coordinates on machine $i$ are equal to \eqref{eqn:cp:coord-on-i}. There are $(p^*)^{(t-1)/t} = O(p^{(t-1)/t})$ such machines. Since the machine has $n/p$ elements initially, it sends
\myeqn{
    O \left(
        \fr{n}{p} \cdot p^{\fr{t-1}{t}}
    \right)
    =
    O \left(
        \fr{n}{p^{1/t}}
    \right)
    \nn
}
elements.

\vgap

Next, we analyze how many elements are received by a machine. If a machine is not in the grid, it receives nothing. Now, consider a machine with coordinates $(x_1, x_2, ..., x_t)$. For each $i \in [t]$, the machine receives all the elements of $R_i$ whose coordinates $y$ satisfy
\myeqn{
    x_i = 1 + \left(y \mod \lf p^{1/t} \rf \right). \nn
}
There can be at most $1 + |R_i| / \lf p^{1/t} \rf = O(|R_i| / p^{1/t})$ such elements. Therefore, the total number of elements received by the machine is bounded by
\myeqn{
    O\left(
        \sum_{i=1}^t
        \fr{|R_i|}{p^{1/t}}
    \right)
    =
    O\left(
        \fr{n}{p^{1/t}}
    \right). \nn
}

\extraspacing {\bf A Lower Bound.} We prove that any algorithm solving the CP problem must incur a load of $\Omega(n / p^{1/t})$ in the worst case, regardless of how many rounds it performs.

\vgap

As a hard input, we set the size of each $S_i$ to be $n/t$, for all $i \in [t]$. The CP (Cartesian product) has a total size of $(n/t)^t$. Suppose that an algorithm can guarantee a load of $L$.  This means that, throughout the algorithm, each machine can see at most $L + \fr{n}{p}$ distinct elements in total and, hence, can produce at most $(L + \fr{n}{p})^t$ tuples in the CP. This yields:
\myeqn{
    t \cdot \left(L + \fr{n}{p} \right)^t
    &\ge&
    \left(
        \fr{n}{t}
    \right)^t. \nn
}
Solving the inequality gives $L = \Omega(n / p^{1/t})$.

\vgap

As a corollary, if each round is allowed to incur a load of $\lambda$, then $\Omega(\fr{n / p^{1/t}}{\lambda})$ rounds are needed to solve the LP problem.


%\section{Remark} ksv10

\bibliographystyle{plainurl}% the mandatory bibstyle
\bibliography{ref}

\appendix

\section*{Appendix: Chernoff Bounds}

%\begin{lemma} \label{lmm::chernoff}
    Let $X_1, X_2, ..., X_f$ be independent Bernoulli random variables that are identically distributed. Define $X = \sum_{i \in [f]} X_i$. For any $\gamma \geq 2$, it holds that
    \myeqn{
        \Pr[X \geq \gamma \cdot \expt[X]] &\leq&  \exp(-\gamma \cdot \expt[X] / 6). \label{eqn:chernoff1}
    }
    For any $0 < \gamma < 1$, it holds that
    \myeqn{
        \Pr[X \le (1 - \gamma) \expt[X]] &\leq&  \exp(- \gamma^2 \cdot \expt[X] / 3). \label{eqn:chernoff2}
    }
%\end{lemma}
Proofs for \eqref{eqn:chernoff1} and \eqref{eqn:chernoff2} can be found in \cite{stl12}.

\end{document}
